### 2025-12-18 articles/arxiv/paper.tex
Updates from Overleaf

    On May 17, 2025, the Chicago Cubs hosted the Chicago White Sox in the second game of their Crosstown Series. In the bottom of the first inning, White Sox pitcher Sean Burke faced Cubs batter Michael Busch with two outs and Cubs runner Kyle Tucker on first base. Before the 1-1 pitch, Burke pivoted and threw to first base behind Tucker, who retreated safely before a tag. Tucker had 10 stolen bases in 10 attempts on the young season, leading Major League Baseball among base stealers who had not been caught \parencite{fangraphs_major_2025}. On the White Sox broadcast, color commentator Steve Stone explained to the audience, ``The only problem is now there's two throws over there, and Tucker's gonna get a much bigger lead.'' On the next pitch, Tucker stole his 11$^{th}$ base of the season, the first Burke had allowed. ``I don't think anybody in the room could have thrown him out on that play,'' said Stone.

@@

    {\color{blue}{Section \ref{sec:sg-model} uses a two-payer zero-sum sequential game to model the evolving interaction between the runner and the pitcher.  Our model captures all the vital aspects of the play while retaining analytical and computational tractability. Our model is closely related to the class of stochastic/dynamic Stackelberg games \parencite{goktas2022zero,li2017review,lopez2022stationary,vasal2020stochastic,vorobeychik2012computing}, which combines the elements of Stackelberg and stochastic games. The main difference between our infinite-horizon model and the literature on stochastic Stackelberg games is that we do not discount the rewards. The rationale for this modeling choice is that the rewards in our model correspond to the runs scored on a play, and runs scored on later plays count the same as those scored on earlier ones. We establish a maximin Bellman equation for our sequential game and propose a variant of value iteration and policy iteration to determine the optimal strategies for runner and the pitcher.}}

@@

      The most striking observation is that the optimal lead distance is significantly longer with two prior disengagements (14.8--16.1 feet) than in other states. These recommended lead distances are near the edge of observed lead distances in the dataset but not outside of the observed range. Overall, the optimal runner policy is more aggressive than observed behavior. Recall from Section \ref{sec:results-multilevel-model} that the average lead distance after zero, one, and two prior disengagements was 9.6 feet, 10.3 feet, and 11.0 feet, respectively.

@@

      Under the game-theoretic equilibrium, pitchers would be more aggressive with attempting pickoffs when runners take long leads. Based on observed lead distances in 2023, we estimate that a pickoff attempt would reduce run expectancy relative to a pitch in 8.3\% of opportunities with only first base occupied. When a pickoff attempt would have been recommended, pitchers attempted pickoffs 10.8\% of the time---this would be 100\% in the two-player model. When a pickoff attempt would not have been recommended, pitchers attempted pickoffs 5.0\% of the time---this would be 0\% in the two-player model. Overall, pitchers attempted pickoffs on 5.8\% of opportunities, less than the recommended 8.3\%.

@@

    Given observed lead distance behavior, offenses scored 0.516 runs per inning in 2023. In other words, if we model the inning as a Markov reward process using empirical transition probabilities between states, the value of the start-of-inning state is 0.516 runs. Under the two-player game, the value of the start-of-inning state is 0.520 runs, which corresponds to approximately 5 more runs over a full season of 162 games of nine innings each. Therefore, offenses stand to benefit more in the transition from observed behavior to the game-theoretic equilibrium. Under the one-player MDP, the value of the start-of-inning state is 0.528 runs, which corresponds to approximately 17 more runs over a full season, relative to observed behavior. Naturally, scoring expectation is higher in the one-player MDP because the runner exploits the sub-optimal strategy of the pitcher. To add 17 expected runs in free agency would cost roughly \$17 million \parencite{clemens_what_2021}, which speaks to a significant (but not outlandish) impact. This value may have increased since the last study we found to have estimated it.

@@

    We go a step further and consider the opportunity for runners to exploit the observed suboptimal behavior by pitchers. We consider a one-player MDP model in which the pitcher's agency in the pickoff attempt decision is replaced by a stochastic process, modeled on empirical data as a function of the runner's lead distance.  This framework leads to the most actionable recommendation for runners under the current conditions of pitcher behavior. We estimate that the average team could expect to increase their expected offensive output by approximately 17 runs over a full season by implementing the optimal lead distance policy. Based on these results, we propose the Two-Foot Rule: Extend your lead by two feet after each pickoff attempt. This easily implemented recommendation is a good approximation to the optimal lead distance policy across a wide array of contexts (outs and ball-strike count) and pitcher/runner skills.

>>>>

    On May 17, 2025, the Chicago Cubs hosted the Chicago White Sox in the second game of their Crosstown Series. In the bottom of the first inning, White Sox pitcher Sean Burke faced Cubs batter Michael Busch with two outs and Cubs runner Kyle Tucker on first base. Before the 1-1 pitch, Burke pivoted and threw to first base behind Tucker, who retreated safely before a tag. Tucker had 10 stolen bases in 10 attempts on the young season, leading Major League Baseball (MLB) among base stealers who had not been caught \parencite{fangraphs_major_2025}. On the White Sox broadcast, color commentator Steve Stone explained to the audience, ``The only problem is now there's two throws over there, and Tucker's gonna get a much bigger lead.'' On the next pitch, Tucker stole his 11$^{th}$ base of the season, the first Burke had allowed. ``I don't think anybody in the room could have thrown him out on that play,'' said Stone.

@@

    {\color{blue}{Section \ref{sec:sg-model} uses a two-payer zero-sum sequential game to model the evolving interaction between the runner and the pitcher.  Our model captures all the vital aspects of the play while retaining analytical and computational tractability. Our model is closely related to the class of stochastic/dynamic Stackelberg games \parencite{goktas2022zero,li2017review,lopez2022stationary,vasal2020stochastic,vorobeychik2012computing}, which combines the elements of Stackelberg and stochastic games. The main difference between our infinite-horizon model and the literature on stochastic Stackelberg games is that we do not discount the rewards. The rationale for this modeling choice is that we choose our optimization objective to be expected runs in the inning, which by definition includes no discount. We establish a maximin Bellman equation for our sequential game and propose a variant of value iteration and policy iteration to determine the optimal strategies for runner and the pitcher.}}

@@

      The most striking observation is that the optimal lead distance is significantly longer with two prior disengagements (14.8--15.8 feet) than in other states. These recommended lead distances are near the edge of observed lead distances in the dataset but not outside of the observed range. Overall, the optimal runner policy is more aggressive than observed behavior. Recall from Section \ref{sec:results-multilevel-model} that the average lead distance after zero, one, and two prior disengagements was 9.6 feet, 10.3 feet, and 11.0 feet, respectively.

@@

      Under the game-theoretic equilibrium, pitchers would be more aggressive with attempting pickoffs when runners take long leads. Based on observed lead distances in 2023, we estimate that a pickoff attempt would reduce run expectancy relative to a pitch in 11.4\% of opportunities with only first base occupied. When a pickoff attempt would have been recommended, pitchers attempted pickoffs 10.4\% of the time---this would be 100\% in the two-player model. When a pickoff attempt would not have been recommended, pitchers attempted pickoffs 4.9\% of the time---this would be 0\% in the two-player model. Overall, pitchers attempted pickoffs on 5.6\% of opportunities, less than the recommended 11.4\%.

@@

    Given observed lead distance behavior, offenses scored 0.516 runs per inning in 2023. In other words, if we model the inning as a Markov reward process using empirical transition probabilities between states, the value of the start-of-inning state is 0.516 runs. Under the two-player game, the value of the start-of-inning state is 0.520 runs, which corresponds to approximately 6 more runs over a full season of 162 games of nine innings each. Therefore, offenses stand to benefit more in the transition from observed behavior to the game-theoretic equilibrium. Under the one-player MDP, the value of the start-of-inning state is 0.524 runs, which corresponds to approximately 12 more runs over a full season, relative to observed behavior. Naturally, scoring expectation is higher in the one-player MDP because the runner exploits the sub-optimal strategy of the pitcher. To add 12 expected runs in free agency would cost roughly \$12 million \parencite{clemens_what_2021}, which speaks to a significant (but not outlandish) impact. This value may have increased since the last study we found to have estimated it.

@@

    We go a step further and consider the opportunity for runners to exploit the observed suboptimal behavior by pitchers. We consider a one-player MDP model in which the pitcher's agency in the pickoff attempt decision is replaced by a stochastic process, modeled on empirical data as a function of the runner's lead distance.  This framework leads to the most actionable recommendation for runners under the current conditions of pitcher behavior. We estimate that the average team could expect to increase their expected offensive output by approximately 12 runs over a full season by implementing the optimal lead distance policy. Based on these results, we propose the Two-Foot Rule: Extend your lead by two feet after each pickoff attempt. This easily implemented recommendation is a good approximation to the optimal lead distance policy across a wide array of contexts (outs and ball-strike count) and pitcher/runner skills.

### 2025-12-18 output/tables/lead_by_players.tex
Updates from Overleaf

\begin{tabular}{cc|rrr}
\multicolumn{2}{c|}{Skill Percentile} & \multicolumn{3}{c}{Disengagements}\\
Battery & Runner & 0 & 1 & 2 \\
% latex table generated in R 4.5.0 by xtable 1.8-4 package
% Sat Aug  2 17:11:21 2025
  \hline
10th & 10th & 10.5 & 11.8 & 14.1 \\ 
  10th & 50th & 11.2 & 12.3 & 14.3 \\ 
  10th & 90th & 11.8 & 12.7 & 14.5 \\ 
   \hline
50th & 10th & 9.4 & 10.9 & 13.7 \\ 
  50th & 50th & 9.9 & 11.4 & 13.8 \\ 
  50th & 90th & 10.6 & 11.9 & 14.0 \\ 
   \hline
90th & 10th & 8.1 & 9.8 & 12.5 \\ 
  90th & 50th & 8.6 & 10.1 & 12.6 \\ 
  90th & 90th & 9.2 & 10.5 & 12.7 \\ 
\end{tabular}

>>>>

\begin{tabular}{cc|rrr}
\multicolumn{2}{c|}{Skill Percentile} & \multicolumn{3}{c}{Disengagements}\\
Battery & Runner & 0 & 1 & 2 \\
% latex table generated in R 4.5.0 by xtable 1.8-4 package
% Thu Dec 18 00:04:13 2025
  \hline
10th & 10th & 10.4 & 11.7 & 14.1 \\ 
  10th & 50th & 11.0 & 12.1 & 14.2 \\ 
  10th & 90th & 11.9 & 12.8 & 14.6 \\ 
   \hline
50th & 10th & 9.4 & 11.1 & 13.8 \\ 
  50th & 50th & 10.0 & 11.4 & 13.9 \\ 
  50th & 90th & 11.0 & 12.2 & 14.2 \\ 
   \hline
90th & 10th & 8.2 & 9.9 & 12.6 \\ 
  90th & 50th & 8.7 & 10.2 & 12.6 \\ 
  90th & 90th & 9.7 & 10.9 & 12.9 \\ 
\end{tabular}
